---
title: "CodeBook for the tidy dataset"
output: html_document
---

# Data Source

The dataset used for this project comes from the "Human Activity Recognition Using Smartphones Data Set" originally available from this site: http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

# Getting the Data and Overview of Transformations with Comments
## Preliminaries
```{r message=FALSE, results='hide'}
packages <- c("data.table", "reshape2")
sapply(packages, require, character.only=TRUE, quietly=TRUE)

path <- getwd()
path
```
## Download the file and put in folder
```{r message=FALSE, results='hide'}
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
f <- "Dataset.zip"
if (!file.exists(path)) {dir.create(path)}
download.file(url, file.path(path, f))
```

## Unzip the file
```{r message=FALSE, results='hide'}
unzip("./Dataset.zip")
```

## Put unzipped files into UCI HAR Dataset folder
```{r message=FALSE, results='hide'}
pathIn <- file.path(path, "UCI HAR Dataset")
```

## Read the subject files
```{r message=FALSE, results='hide'}
SubjectTrain <- read.table(file.path(pathIn, "train", "subject_train.txt"))
SubjectTest  <- read.table(file.path(pathIn, "test" , "subject_test.txt" ))
```

## Read the activity files
```{r message=FALSE, results='hide'}
ActivityTrain <- read.table(file.path(pathIn, "train", "Y_train.txt"))
ActivityTest  <- read.table(file.path(pathIn, "test" , "Y_test.txt" ))
```

## Read the features files, transformed to a data.table object using the fileToDataTable helper function
```{r message=FALSE, results='hide'}
fileToDataTable <- function (f) {
  df <- read.table(f)
  dtbl <- data.table(df)
}
Train <- fileToDataTable(file.path(pathIn, "train", "X_train.txt"))
Test  <- fileToDataTable(file.path(pathIn, "test" , "X_test.txt" ))
```

## Merge the training and testing sets
### Concatenate the data tables 
```{r message=FALSE, results='hide'}
Subject <- rbind(SubjectTrain, SubjectTest)
setnames(Subject, "V1", "subject")
Activity <- rbind(ActivityTrain, ActivityTest)
setnames(Activity, "V1", "activity_num")
dtbl <- rbind(Train, Test)
```

### Merge the columns and make sure dtbl is a data.table object (in order to set the key)
```{r message=FALSE, results='hide'}
Subject <- cbind(Subject, Activity)
dtbl <- cbind(Subject, dtbl)
dtbl <- as.data.table(dtbl)
```

### Set the key for the data for use later
```{r message=FALSE, results='hide'}
setkey(dtbl, subject, activity_num)
```

## Extract mean and std
### Read the features.txt file, which tells us which varibles in dtbl are measurements for the mean and std
```{r message=FALSE, results='hide'}
Features <- fread(file.path(pathIn, "features.txt"))
setnames(Features, names(Features), c("feat_num", "feat_name"))
```
### Subset the measurements for mean and standard deviation using grepl
```{r message=FALSE, results='hide'}
Features <- Features[grepl("mean\\(\\)|std\\(\\)", feat_name)]
```
### Convert the column numbers to a vector of variable names which match the columns of dtbl
```{r message=FALSE, results='hide'}
Features$feat_code <- Features[, paste0("V", feat_num)]
```
### Subset these variables from above using the variable names 
```{r message=FALSE, results='hide'}
select <- c(key(dtbl), Features$feat_code)
dtbl <- dtbl[, select, with=FALSE]
```

## Descriptive Activity Names
### Read the activity_labels.txt file which can be used to add descriptive names to label the activities
```{r message=FALSE, results='hide'}
ActivityNames <- fread(file.path(pathIn, "activity_labels.txt"))
setnames(ActivityNames, names(ActivityNames), c("activity_num", "activity_name"))
```

### Merge the activity labels 
```{r message=FALSE, results='hide'}
dtbl <- merge(dtbl, ActivityNames, by="activity_num", all.x=TRUE)
```

### Add activity_name as a key for the data for later access ease 
```{r message=FALSE, results='hide'}
setkey(dtbl, subject, activity_num, activity_name)
```

### Melt the data.table in order to reshape it into a short and wide format
```{r message=FALSE, results='hide'}
dtbl <- data.table(melt(dtbl, key(dtbl), variable.name="feat_code"))
```

### Merge the data
```{r message=FALSE, results='hide'}
dtbl <- merge(dtbl, Features[, list(feat_num, feat_code, feat_name)], by="feat_code", all.x=TRUE)
```

### Create new variable called activity, which is equivalent to activity_name, as a factor class. Also create a new variable feature in dtbl, which is equivalent to feat_name, as a factor class
```{r message=FALSE, results='hide'}
dtbl$activity <- factor(dtbl$activity_name)
dtbl$feature <- factor(dtbl$feat_name)
```

### Helper function grepthis helps separate features from feature_name for labeling purposes, creating the variables for the tidy data set.  The variables are described later in this CodeBook.
```{r message=FALSE, results='hide'}
grepthis <- function (regex) {
  grepl(regex, dtbl$feature)
}
```

### Using the helper function on features with 2 categories
```{r message=FALSE, results='hide'}
n <- 2
y <- matrix(seq(1, n), nrow=n)
x <- matrix(c(grepthis("^t"), grepthis("^f")), ncol=nrow(y))
dtbl$featDomain <- factor(x %*% y, labels=c("Time", "Freq"))
x <- matrix(c(grepthis("Acc"), grepthis("Gyro")), ncol=nrow(y))
dtbl$featInstrument <- factor(x %*% y, labels=c("Accelerometer", "Gyroscope"))
x <- matrix(c(grepthis("BodyAcc"), grepthis("GravityAcc")), ncol=nrow(y))
dtbl$featAcceleration <- factor(x %*% y, labels=c(NA, "Body", "Gravity"))
x <- matrix(c(grepthis("mean()"), grepthis("std()")), ncol=nrow(y))
dtbl$featVariable <- factor(x %*% y, labels=c("Mean", "SD"))
```

### using the helper function on features with 1 category
```{r message=FALSE, results='hide'}
dtbl$featJerk <- factor(grepthis("Jerk"), labels=c(NA, "Jerk"))
dtbl$featMagnitude <- factor(grepthis("Mag"), labels=c(NA, "Magnitude"))
```

### Using the helper function on features with 3 categories
```{r message=FALSE, results='hide'}
n <- 3
y <- matrix(seq(1, n), nrow=n)
x <- matrix(c(grepthis("-X"), grepthis("-Y"), grepthis("-Z")), ncol=nrow(y))
dtbl$featAxis <- factor(x %*% y, labels=c(NA, "X", "Y", "Z"))
```

### Create a tidy data set, which features the average of each variable for every activity and subject
```{r message=FALSE, results='hide'}
setkey(dtbl, subject, activity, featDomain, featAcceleration, featInstrument, featJerk, featMagnitude, featVariable, featAxis)
tidy_data <- dtbl[, list(count = .N, average = mean(value)), by=key(dtbl)]
```

# Variables and Descriptions
Variable | Description
---------|-------------
subject  | ID's the subject who performed the activity for each window sample. Ranges from 1 to 30
activity | the name of the activity
featDomain | a feature: time domain signal or frequency domain signal (Time or Freq)
featAcceleration | a feature: acceleration signal (Body or Gravity)
featInstument | a feature: the measuring instument (Accelerometer or Gyroscope)
featJerk |  a feature: Jerk signal 
featMagnitude | a feature: magnitude of the signals
featVariable | a feature: variable (mean or standard deveiation)
featAxis | a feature: 3 axial signals in the X, Y, and Z directions
count | a feature: count of the data points used to compute the average
average | a feature: the average of each variable for each subject and each activity

# Structure of the Dataset
```{r}
str(tidy_data)
```


